---
title: "DrugTestAnalysisMixedModels"
author: "Duc-Anh Nguyen"
date: "2025-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


bloodpress is a fictional dataset
```{r}
library(ggplot2)
#LadenderDatenundUberblickuberDatensatz:
load("bloodpress.RData")
head(bloodpress)
dim(bloodpress)
skimr::skim(bloodpress)
colnames(bloodpress)

table(bloodpress$Person)
table(bloodpress$Dose)
table(bloodpress$SBP)

bloodpress |> ggplot(aes(Dose, SBP, col = as.factor(Person))) + geom_point() + geom_line() + facet_wrap(~Gender) + ggtitle("Blood pressures for different doses in 30 people") + ylab("Blood pressure in mmHg") + xlab("Dose in mg")+ scale_color_discrete(name = "30 people (test subjects)")


```
There is a trend for both *MOST* men and women: the higher the dose the lower the blood pressure (negative correlation)


```{r}
library(lme4)
interaction_mixed_lm <- lmer(formula = SBP ~ Gender*Dose + (1|Person), data = bloodpress, REML = FALSE)
summary(interaction_mixed_lm)
std_dev <- as.data.frame(VarCorr(interaction_mixed_lm))
(ICC = (std_dev$sdcor[1]^2)/sum(std_dev$sdcor^2))


```
So the percentage of variance explained by the random intercept, that is not explained by the fixed effects, is around 50%


Now we want to check whether the interaction between gender and dose is necessary by a likelihood ratio test, with significant level of 0.05
Formula: H~0~: $β_{interaction} = 0 $ vs. H1:$β_{interaction} \neq 0 $

Remark: to carry this likelhood ratio test we will want to fit the both model with ML and not REML!
```{r}
mixed_lm <- lmer(formula = SBP~Gender + Dose + (1|Person), data = bloodpress, REML = FALSE)
anova(interaction_mixed_lm, mixed_lm)
```
With that p value we could not reject the H0, so we have to stay with that that the interaction makes no significant difference! 

Now we want to test if the random intercept makes sense. We will test the random intercept : 
$\tau_0 = 0 $ vs. H1:$\tau_0 \geq 0 $
```{r}
lmodel <- lm(SBP~ Gender*Dose, data = bloodpress)
# approximated test based on ML
anova(interaction_mixed_lm, lmodel)


# the exact test based on REML
RLRsim::exactRLRT(m = update(interaction_mixed_lm, REML = TRUE))

```
Both of the test lead to a p value hinting that we could reject the H0, which tells us that we should keep the random intercept in the model.


From the plots in the beginning we can see that each person reacts to the rise of dose differently, so it might make sense to also include a random slope for the doses
```{r}
full_model <- lmer(SBP~Gender*Dose + (1+Dose|Person), data = bloodpress, REML = FALSE)
anova(full_model, interaction_mixed_lm)


```
The test shows that it makes sense to include the random slopes, as it does make a difference 


Let us check also the model with random slope and no random intercept, with exact test and conditional AIC.
Conditional AIC is suitable for when models are having the same fixed effects but different random effects
```{r}
# exact test
randomslope_model <- lmer(
  formula = SBP ~ Gender * Dose + (-1 + Dose | Person),
  data = bloodpress,
  REML = FALSE
)
RLRsim::exactRLRT(m = full_model, mA = randomslope_model, m0 = interaction_mixed_lm)

# cAIC
# install.packages("cAIC4")
library(cAIC4)
cAIC(lmodel)
cAIC(interaction_mixed_lm)
cAIC(randomslope_model)
cAIC(full_model)
```
